---
title: "Report Excercise 3: Cleaning Data from elevated CO2 Experiments"
author: "Patricia Gribi"
date: "2023-03-02"
output: html_document
---

In a first step the data from the paper “Faster Decomposition Under Increased Atmospheric CO2 Limits Soil Carbon Storage.” was cleaned and converted into a CSV file that can be read into r.

```{r}

library(tidyverse)

soil_organic_carbon <- read_csv("./data/03ex_tidy_data.csv") 

```

As a next step the different experiments were grouped. Furthermore the log-response ratio for each experiment was calculated. 

Describe more, insert Einheiten, Tabelle formatieren, anderer Variablen-Name

Wegen Einrückung fragen

What is ambient CO2?

```{r}

soil_organic_carbon_summary <- soil_organic_carbon |>  
  group_by(Experiment) |> 
  summarise(across(ends_with("CO2_mean"), mean),
            log_response_ratio = log(increased_CO2_mean/ambient_CO2_mean, base= exp(1)))

knitr::kable(soil_organic_carbon_summary)

```



```{r}

soil_organic_carbon |>
  mutate(Experiment_Phase = ifelse(`Time_(years)` <= 3, "early phase", 
                            ifelse(`Time_(years)` >= 6, "mid-phase","late phase")))|> # the Experiments are converted into 3 phases
  group_by(Experiment_Phase) |> 
  summarise(across(ends_with("CO2_mean"), mean),
            log_response_ratio = log(increased_CO2_mean/ambient_CO2_mean, base= exp(1)))


```

